---
title: "ModelBuilding_Predictors"
author: "Serena Chang"
date: "2026-06-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started with our models

*What is a model?*

According to [Jackson et al., 2000](https://doi.org/10.1641/0006-3568(2000)050[0694:AITTPO]2.0.CO;2), a *model* is a representation of a thing, idea or a condition. In my opinion this definition is terribly vague, but that doesn't make it less true. A model can be simple (think a picture or a statement) or complex (think advanced mathematical equations). *Modeling* is the process by which a hypothesis evolves from an idea into a conceptual model into a quantitative model.

An idea -> data collection -> model development -> new understanding -> revised hypothesis -> more data -> more models -> and on and on and on...

Models should be detailed enough to encompass the ecological structure and processes of interest but simple enough to be understood and not overfit.

For this study, we started with a question from which we developed a conceptual model based on our understanding of our model system: Lyme disease in the western United States. 

Now we will begin attaching data to our question to make predictions. We will be building an *empirical model*. Empirical models allow us as researchers to identify relationships between variables using real-world data. As such, it is important for us to identify relevant parameters for our models based on our study system and hypothesis as well as reliable methods/sources by which we can collect/extract that data.


####Install packages
Install packages, if needed. Edit this chunk to include previously uninstalled packages and remove the # symbol.
```{r cars}
#install.packages(c( "tidyverse", "janitor", "lubridate", "sf", "terra", "geodata", "elevatr", "glmmTMB" ))
```

The chunk above is used to install packages hosted on CRAN directly. For packages that need to be sourced directly from GitHub repository, install them with the code as follows:
```{r}
#remotes::install_github("rspatial/luna")
```

####Load packages
```{r pressure, echo=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(sf)
library(terra)
library(geodata)
library(elevatr)
library(glmmTMB)
```

#####Defining our geographic area

Our study takes place in the San Francisco Bay Area. The northern most study area is in Sonoma County at Pepperwood Preserve, the eastern most study area in Contra Costa County as Black Diamond Mines Regional Preserve, the southern most study area in Santa Clara County at Uvas Canyon Regional Park. Based on these points and the Pacific Coastline as the western bound, define the geographic area of our study. The area should have a slight buffer around the furthest extents of our study sites. 

On Tuesday, you did research to identify an appropriate bounding box for our study area.

Defined here:
```{r}
# Define the geographic area (bounding box) by defining the objects below (remove the "#" at the front and replace "Min/Max long/lat for our study area" with the associated coordinates.)

#CRS WGS 84 UTM Zone 13

lon_min <- -122.89
lon_max <- -121.74
lat_min <- 37.2658
lat_max <- 38.6624
```

Study area based on above definition
```{r}
area <- c(lat_min, lat_max, lon_min, lon_max)
```

However, we can also define our study area by manipulating our response datasets and creating a buffer around the study sites in R, which we will do together now. 

First, we will want to take a look at some of our response datasets and get them cleaned up so we can work with them effectively.


### Import and clean site-level and visit-level ecological response data ###

As with all scripts, the first thing we need to do to setup our workspace is install and load all relevant packages. All relevant packages were included in the above chunks, so if you have not loaded them into your library, go ahead and do that now. 

Once you've loaded all relevant packages, you will want to import and take a look at the response data we will be working with today. Associated csv files are available in our shared drive: [Sites](https://drive.google.com/file/d/1sY2M3p13s2IY3jzU-Ctou5wNXWAq_hzK/view?usp=drive_link) & [Visits](https://drive.google.com/file/d/1Wg3GkkfqWnNzLsNGCW9DZLAG2xE1Ysze/view?usp=drive_link).

```{r}
sites_raw <- read_csv("data/Puma_data - Sites_Puma.csv")
visits_raw <- read_csv("data/Puma_data - Visits_Puma_Counts.csv")
```

##### Inspect the data #####

```{r}
glimpse(sites_raw)
glimpse(visits_raw)

summary(sites_raw)
summary(visits_raw)
```

**How are the site-level and visit-level tables different? Why might be need both?**
The site-level table keeps track of information associated with the study site that doesn't change with each visit. The visit-level table includes information about our response variable collections at each site on different dates.

### Data Cleaning ###
Data cleaning allows us to perform checks on our data. Is the dataset accurate and usable for analysis? Raw data often contains errors, including but not limited to missing values, incorrect formats, duplicates, or inconsistent entries. These errors can give us a hard time down the line.

Using the mutate and filter function, clean up the site-level and visit-level datasets to prepare for model building. Consider what each row and column represents to think about what their uses might be and what datatype will be best for that use. Drop any empty rows and think about how you might handle missing data. Why might be filter out observations with missing data is some rows versus others?

```{r}
sites_clean <- sites_raw %>%
  mutate(
    site_id = as.character(site_id),
    lon = as.numeric(longitude),
    lat = as.numeric(latitude),
    puma_factor = as.factor(puma_factor) # we want them as factors so that they are actual levels
    )%>%
  filter(
!is.na(site_id),
!is.na(lon),
!is.na(lat),
between(lon, -180, 180),
between(lat, -90, 90)
)%>%
distinct(site_id, .keep_all=TRUE)


visits_clean <- visits_raw %>%
  mutate(
    site_id = as.character(site_id),
    site = as.character(site),
    drag_num = as.integer(drag_num),
    collection_date = mdy(collection_date),
    puma_factor = as.factor(puma_factor),
    observer_count = as.integer(observer_count),
    area_surveyed_m2 = as.numeric(area_sureveyed_m2),
    count_IPAC_L = as.integer(count_IPAC_L),
    count_IPAC_N = as.integer(count_IPAC_N),
    count_IPAC_A_F = as.integer(count_IPAC_A_F),
    count_IPAC_A_M = as.integer(count_IPAC_A_M),
    don = count_IPAC_N / area_sureveyed_m2, # this is how to calculate density of nymphs
    dol = count_IPAC_L / area_sureveyed_m2,
    doaf = count_IPAC_A_F / area_sureveyed_m2,
    doam = count_IPAC_A_M / area_sureveyed_m2
  ) %>%
  filter(
    !is.na(site_id)
  )

#for visits_clean also add a column for density of nymphs called "don." Which count column should you use?

glimpse(sites_clean)
glimpse(visits_clean)
```

After cleaning your datasets, make sure that each visit has a corresponding site. 

```{r}
anti_join(visits_clean, sites_clean, by="site_id")
```
If this line of code returns zero rows, your site IDs match!


Using the chunk below, create a table showing how many visits each site has and answer the following question: Do all sites have the same number of visits? If not, which sites are missing visits?

No, not all sites have the same number of visits. RCDO, SAN, and UVA are missing visits.

```{r}
visits_clean %>%
  count(site_id)
```


### Site-Level Density Data ###

Summarize the two visits to consider site-level density (we might consider each drag effort at each site as independent observations, but we need to understand it summarized to know what's best for our analysis).
```{r}

site_don <- visits_clean %>%
  group_by(site_id)%>%
  summarise(
    n_visits = n(),
    total_nymphs = sum(count_IPAC_N, na.rm = TRUE),
    total_area_surveyed_m2 = sum(area_sureveyed_m2, na.rm=TRUE),
    pooled_density = total_nymphs / total_area_surveyed_m2,
    density_100m2 = pooled_density * 100,
    mean_visit_density = mean(don, na.rm=TRUE),
    mean_visit_density_100m2 = mean_visit_density * 100,
    sd_visit_density = sd(don, na.rm=TRUE),
    .groups = "drop"
  )

glimpse(site_don)
summary(site_don)
```

Use a left_join to incorporate site coordinates into our density data and inspect the result.
```{r}
site_don <- site_don %>%
  left_join(sites_clean, by="site_id")

glimpse(site_don)
```

Do you notice a difference between pooled_density and mean_visit_density? When are they different?

Yes, there is a difference between pooled_density and mean_visit_density. pooled_density and mean_visit_density are different when the total area surveyed are different.

### Converting our Sites into Spatial Objects ###

In the chunk below, convert the sites to an sf object named "sites_sf" using st_as_sf().

```{r}

sites_sf <- st_as_sf(site_don, #update with your df 
                     coords= c("lon", "lat"), #vector for your coordinates
                     crs = 4326, remove=FALSE)

```

EPSG:4326 is a spherical model of earth represented in decimal degrees. We are going to want to convert our spatial object into a coordinate system measured in meters for our analysis. We will be using EPSG:3857.

**Why is it useful to use a projected coordinate system in meters rather than latitude/longitude?**

It is easier to work with that system because it is more consistent and makes it easier to do our calculations.

Define an object as "crs_model" with the value "EPSG:3310" and reproject sites_sf using st_transform()
```{r}
crs_model <- "EPSG:3310" # crs should be chosen in meters based on data

sites_proj <- st_transform(sites_sf, crs = crs_model)

```


With our study sites transformed, we will now create a a buffer around our sites to act as our study area or aoi (area of interest).

```{r}
aoi <- sites_proj %>%
  st_union()%>%
  st_buffer(50000)%>%
  st_as_sf()
```

This creates a 50 km buffer around our sites. Does this seem like an appropriate buffer for our study? If not, adjust accordingly.


### Loading Predictors using APIs ###

We will work on loading environmental predictors, focusing on climate, elevation, and slope.

#### Preditor 1: Climate using WorldClim ####

WorldClim is a database of global weather and climate data. They have datasets that look at annual trends of bioclimatic variables. You can find the codes for these variables [here](https://www.worldclim.org/data/bioclim.html)

When we load these datasets, we are going to save them in a new folder within the data foloder of our project.

```{r}
dir.create("data/predictors/worldclim", recursive=TRUE, showWarnings = FALSE)
```

Bioclimatic data is available at multiple resolutions through WorldClim. We are going to attempt to load the data at 2.5 resolution, but if time and bandwidth become a limiting factor we can load it at 10 resolution.

```{r}
bio <- geodata::worldclim_global(
  var="bio",
  res = 2.5,
  path = "data/predictors/worldclim"
)
```

We are going to keep four climate predictors.

```{r}
bio_worlclim <- bio[[c(1, 2, 4, 12)]]
names(bio_worlclim) <- c("annual_mean_temp", "mean_diurnal_range", "temp_seasonality", "annual_precipitation")
plot(bio_worlclim)
```

#### Predictor 2: Elevation ####
We will now download elevation data *near our sites* using elevatr.

```{r}
aoi_II <- st_transform(aoi, 4326)

elev_raw <- elevatr::get_elev_raster(
  locations = aoi_II,
  z=8,
  clip="bbox"
)

elev <- terra::rast(elev_raw)
names(elev) <- "elevation_m"
```

#### Predictor 3: Slope ####
Calculate slope from the elevation

```{r}
slope <-  terra::terrain(
  elev,
  v="slope",
  unit = "degrees" # units are degrees b/c we are in 4326
)

names(slope) <- "slope_deg"
```

Plot each predictor:

```{r}
bio_worlclim <- bio_worlclim %>%
  crop(vect(aoi_II)) %>%
  mask(vect(aoi_II))


# plots each map for each bio predictor
plot(bio_worlclim[[1]])
plot(st_geometry(sites_sf), add=TRUE)

plot(bio_worlclim[[2]])
plot(st_geometry(sites_sf), add=TRUE)

plot(bio_worlclim[[3]])
plot(st_geometry(sites_sf), add=TRUE)

plot(bio_worlclim[[4]])
plot(st_geometry(sites_sf), add=TRUE)


plot(elev) # shows the elevation of the study area (e.g. mountains)
plot(st_geometry(sites_sf), add=TRUE)

plot(slope)
plot(st_geometry(sites_sf), add=TRUE, col="white")
```

Based on these plots, which predictor seems to vary the most across the studdy area?


### Formatting Our Predictors ###

Because spatial predictors can have different resolutions, projections, and extents, it is important that we get them all to match our response variable projection before we extract any values. This allows us to compare apples to apples instead of apples to oranges. 

Make a template of our response variable projection in raster format:

```{r}
template <- rast(
  ext(vect(aoi)),
  resolution = 1000,
  crs = crs_model
)
```

Project climate and elevation predictors using this template.

```{r}
bio_model <- project(bio_worlclim, template, method="bilinear")
elev_model <- project(elev, template, method="bilinear")
slope_model <- project(slope, template, method="bilinear")

```

Now combine the predictors into a raster stack by creating a vector of the above models. Name the object "predictor_stack"
```{r}
predictor_stack <- c(bio_model, elev_model, slope_model)
```

### Extracting Site-Level Predictors ###

We will now extract the raster values to the associate sites.

```{r}
site_predictors <- terra::extract(
  predictor_stack,
  vect(sites_proj)
)%>%
  as_tibble()%>%
  mutate(site_id = sites_clean$site_id)%>%
  select(site_id, everything(), -ID)

#Inspect the results
```

### Create Final Tables to Use when Modeling ###

Using left_join() create a dataframe that combines your predictor variable values with the site-level density data and another dataframe that combines your predictor variable values with the visit-level data.

```{r}
site_level <- sites_clean %>%
  left_join(site_predictors, by="site_id")

glimpse(site_level)
```

```{r}
visit_level <- visits_clean %>%
  left_join(site_level, by="site_id") %>%
  left_join(site_predictors, by="site_id")

glimpse(visit_level)
```

**Why might visit-level data be better or worse for modeling than the site level average density data? How would you do this process for a dataframe that contains a unique row for each tick collected in the field? How might sample-level data be better or worse for modeling than visit or site-level data?**

Visit-level data might be worse since there are less visits to some sites, because the data might become biased.

###Additional questions ###

1. Why is count not the same thing as density?

Density is the count divided by the area. This is important because the data distribution is going to look different in a count-based model vs a density-based model and for determining what is an appropriate model for your data or model geometry.

2. What would happen if one site had twice as much area surveyed as another site but we ignored survey area?

If one site has twice as much area survey as another site but we ignored the survey area, we would bias our data, making some counts look inflated and others deflated.

3. Why might repeated visits help us understand variation better than a single visit?



4. What is one reason climate might affect the response variable? Elevation? Slope?

Ticks like certain levels of humidity and temperature to sustain their population/transform between life stages, so climate can affect the timing in which they molt.

Elevation could affect the species communities in an area, oxygen/CO2 availibility, humidity levels, etc that can all affect tick populations.

Slope could affect tick populations based on the species community able to utilize that area, and it could affect the samplers' ability to conduct a thorough tick drag.


### Challenge ###

1. Try to fit simple site and visit-level models. Either lm, glm, glmmtmb.

```{r}
visit_level %>%
  lm(formula = don ~ annual_mean_temp.x) %>%
  summary()
# if summary() shows that the signif. code is not *, then you are missing a predictor variable that could further explain the variation in the response variable, don

# the following code, you can see the correlation between x and y
# this can be used for other variables to see what correlates with our response variable
ggplot(data=visit_level, aes(x=annual_mean_temp.x, y=don)) +
  geom_point() +
  geom_smooth(method=lm)
```

y = mx + b


2. Try to standardize the numeric predictors of our resulting site-level and visit-level dataframes that incorporate our predictors. Then refit the models.