---
title: "ModelBuilding_Predictors_Prism"
author: "Jadyn Zapata"
date: "2026-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set-up

This is practice for creating a model and building predictors. This model will use a map from the prism package.

#### Installing packages
```{r install-packages}
# Remove the comment on this code if this package is not downloaded on your device
#install.packages("prism", "gdalUtilities", "raster", "ggplot2", "dplyr", "sp")
```

#### Loading packages
```{r library}
library(prism)
library(gdalUtilities)
library(raster)
library(ggplot2)
library(dplyr)
library(sp)
```

##### Define geographic area

This information was taken from the original Model Building Predictors file.

The geographical area where the study took place is defined with the following variables:
```{r coordinates}
lon_min <- -123.3
lon_max <- -121.8
lat_min <- 37.05
lat_max <- 38.6
```

The study area is defined as:
```{r area}
area <- c(lat_min, lat_max, lon_min, lon_max)
```

#### Setting the temporal extent of the study

The response variable data was collected March 2026 - June 2026. However, this may be too recent of a timeframe, and the package used to retrieve map data may not have been updated for this time.

```{r}
# Set the date range
start <- "2026-03-01"  # yr-mo-d
end <- "2026-06-25"
```

#### Downloading prism data and preparing it for analysis

This section will go into how to retrieve prism data with guidance from the README for prism on the CRAN r-project website and by also following along from the original Model Building Predictors file.

#### Retrieving mean temperature ####

The variable that will be retrieved for this file is tmean, the mean temperature.

First, we need to set the download directory.
```{r}
# This sets the prism archive
prism_set_dl_dir("C:/Users/jadyn/OneDrive/Desktop/R-PINC-LYME PROJECT/LymePINCProj2026/data")
```

Downloading the daily tmean will take too long. Instead, we will download the monthly mean temperatures based on the start and end dates, variables that were initialized earlier.
```{r}
get_prism_monthlys(
  type = "tmean",
  years = 2026,
  mon = c(3, 4, 5, 6),
  resolution = "4km",
  keepZip = FALSE
)
```

Turn the prism files into a raster object.
```{r}
mean_temp_raster <- pd_stack(prism_archive_ls())
```

Save your raster object as a tif file
```{r}
# Make sure the file is saved to the correct location
writeRaster(mean_temp_raster, filename = "mean_temp_prism.tif", format = "GTiff", overwrite = TRUE)
```